---
title: "General Workflow"
author: "Martin Kerschbaumer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The *mbte*-package aims to generalize the workflow regarding the exploratory
data analysis of time-dependent data. The approach is based on the
*[Managing many models](https://speakerdeck.com/hadley/managing-many-models)*-
talk by Hadley Wickham.

# Example workflow

First, the required packages are loaded:
```{r load pkg, message=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(mbte)
```

The *raw_signals*-dataset is used for this demonstration.

```{r show raw dataset, echo=TRUE}
data("raw_signals")
raw_signals
```

The `raw_signals`-dataset is an exclusively fictional dataset, which consists
of 3 columns:

+ *t*: The time-column indicating when the measurement was made.
+ *mv*: The measurement-variable - the measured parameter.
+ *value*: The intensity of the measurement variable.

The goal is to estimate, if trends of different measured parameters are present.

## Creation of the main datastructure

Because `raw_signals` is a tibble, `new_tbl_mbte()` is used to convert it to
a `tbl_mbte` (the main datastructure of the mbte-package). `new_tbl_mbte()`
takes tibbles as first argument or objects, that can be converted to tibbles via
`tibble::as_tibble()`. Additionally, column names for the `time` and the
`value`-columns are passed too.

```{r new_tbl}
# pass as strings
raw_signals <- new_tbl_mbte(raw_signals, time = "t", value = "value")
```

It would be possible to rely on quotation too:
```{r new tbl quotation, eval=FALSE}
raw_signals <- new_tbl_mbte(raw_signals, time = t, value = value)
```

## Nesting of signals

In the following step, the `time`- and `value`- columns get combined into a new
list-column, which is consisting of tibbles (signal-column). A parameter for
grouping must be specified. Since the goal is to find out which measurement-
variables contain trends, the `mv`-column is used for grouping (contains the
names of the measured parameters).

```{r nest}
nested <- mbte_nest_signals(raw_signals, "mv")
nested
```

Also here, the passed variables get quoted. Therefore, the following call is
equivalent to the one above.
```{r use quotation, eval=FALSE}
nested <- mbte_nest_signals(raw_signals, mv)
```

Each subtibble contains the complete signal for the chosen grouping:
```{r show subtibble}
nested$signal[[5]] # print example signal-subtibble
```

## Subsignal extraction

For some specific usecases, it may be required, that the extracted signal-
table gets split into subsignals. This may be necessary if only nonzero
signal-values are analyzed. The following plots show the unextracted
signals. The parts highlighted in blue will be kept (extracted subsignals).
The dashed red line indicates $y = 0$.

```{r plot subsignals, echo=FALSE}
mbte_panel_plot(nested, {
  # use mbte_default_indexer() to show which parts of the signal will be kept
  # (subsignal-extractions)
  ind_tbl <- .u_signals %>% 
    group_by(mv) %>% 
    mutate(ind = {
      list(value %>% mbte_default_indexer() %>% as_tibble())
    }) %>% 
    select(-t, -value) %>% 
    unnest(ind)
  
  ggplot(mapping = aes(t, value)) +
    # highlight subsingals, which will be kept, in blue
    geom_rect(aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
      fill = "#9fb8ff", data = ind_tbl, inherit.aes = FALSE) +
    geom_hline(yintercept = 0, col = "red", linetype = "dashed") +
    geom_path(data = .u_signals, color = "grey10") + # draw signals
    labs(x = "time", y = "signal-values") +
    ggtitle("Raw signals", paste("Page", .n_page, "of", .n_pages)) +
    theme_minimal() +
    theme(axis.text = element_blank(), panel.grid = element_blank())
}, mv, ncol = 7, nrow = 3)
```

The subsignal-extraction is performed by using `mbte_extract_subsignals()`.

```{r extract signals}
extracted <- mbte_extract_subsignals(nested)
extracted
```

The output shows, that the first signal has been split into three subsignals.
The column "signal_nr" has been added to describe the position of the extracted
subsignal within the original signal (e.g. `signal_nr == 2` indicates, that the
corresponding row contains the second subsignal within the orginal signal).

In this example, only subsignals with a length of greater than 11 (arbitrary
threshold) are kept. This should ensure, that signals like the one for
"mv2" (measurement variable 2) are discarded. Filtering can be done using dplyr.
Since `signal` is a list-column, `purrr::map_int()` is used to get the number of
rows of the corresponding signal (which is stored as a tibble).

NOTE: `mbte_reconstruct()` should be called, if functions are invoked, which
may modify the attributes of the original object (in this case
`dplyr::filter()`).

```{r filter subsignals}
filtered_dataset <- extracted %>% 
  filter(purrr::map_int(signal, nrow) > 11) %>% 
  mbte_reconstruct(extracted)
filtered_dataset
```

## Fitting

The remaining rows are analyzed for trends. This is done by fitting an
arbitrary model to the subsignals. The only criteria for the used models is,
that numeric predictions for the original signal must be computable.

Since the general trend is not always known beforehand, `loess()` gets used
to fit a spline to each subsignal using `mbte_fit()`. Additonally, a linear
model gets fit too.

```{r fit}
fitted <- filtered_dataset %>% 
  mbte_fit(
    loess = loess(value ~ t, .signal, span = 0.95),
    lm = lm(value ~ t, .signal)
  )

# resulting table
fitted
```

Users can make use of masked objects for fitting (see documentation of
`mbte_fit()` for details). For example `.signal` does not exist in
the workspace. It gets provided by `mbte_fit()` to simplify fitting. That way,
the user can think of fitting a linear model to a table containing the time-
and value-column (in our example named `t` and `value`). `.signal` is just an
element of the signal-column of the filtered dataset:

```{r}
filtered_dataset$signal[[1]] # example element of `.signal`
```

## Metric computation

In order to compare the fits, an error metric is computed for each fit.
Generally speaking, a trend is present, if the used models are able to
generalize the underlying trend of a signal. Therefore, the lower the resulting
error metric, the higher the likelyhood for an interesting measurement-parameter
worth inversitgating.

For the comparison of the fits, the normalized root mean-squared error is used,
since a error metric is required, which is relative to the range of the
signal-intensities (signal-values).
```{r metrics}
nrmse <- function(obs, pred) {
  sqrt(mean((pred - obs)^2)) / (max(obs) - min(obs))
}

metrics <- mbte_compute_metrics(fitted, nrmse = nrmse(.obs, .pred))
metrics
```

Like for `mbte_fit()`, `mbte_compute_metrics()` also uses masking to simplify
the metric computation. Here `.pred` denote the predicted signal-values from the
model and `.obs` are the observed/original values of the signal (both are
numeric vectors).

The resulting matrics-table can be filterd accordingly using dplyr. A score
gets computed for each signal (identified by the measurement variable and the
signal_nr).

NOTE: In this case, calling `mbte_reconstruct()` is not required, since the
result doesen't have to be of class `tbl_mbte`, which will only be processed
by dplyr and not by the mbte-package directly.
```{r filter metrics}
metrics <- metrics %>% 
  group_by(mv, signal_nr) %>% 
  summarize(score = sum(result)) %>% # compute score (lower is better)
  arrange(-desc(score)) %>% # order by result in ascending order
  ungroup() %>% 
  select(mv, signal_nr, score) # only keep relevant columns
metrics
```

It can be seen, that subsignal nr. `r metrics$signal_nr[1]` of
measurement-variable `r metrics$mv[1]` seems to have the best score. To
investigate the fits further, the `fitted`-table containing the fits is
rearranged (order of rows changed according to the best fits). The
best-performing fits are put at the beginning of the table.

```{r}
fitted <- fitted %>% 
  left_join(metrics, by = c("mv", "signal_nr")) %>% # add "score" column
  arrange(-desc(score)) %>% # reorder rows
  select(-score) %>% # remove column "score"
  mbte_reconstruct(fitted)
fitted
```

## Visualisation of fits

`mbte_panel_plot()` is used to produce the following visualisations of the
fits. Since the `fitted`-table has been rearranged according to the performance
of the corresponding fit, signals containing a clearer trend are drawn first:
```{r plot results}
fit_vis <- mbte_panel_plot(fitted, {
  ggplot(.u_signals, aes(t, value)) +
    geom_point(alpha = 0.75, color = "grey50") +
    geom_path(aes(color = fit, group = paste0(mv, signal_nr, fit)), .u_fits) +
    scale_color_brewer(palette = "Set1") +
    labs(x = "time", y = "signal-values") +
    ggtitle("Fits ordered by metric", paste("Page", .n_page, "of", .n_pages)) +
    theme_minimal() +
    theme(legend.position = "bottom")
}, mv, signal_nr, nrow = 2, ncol = 3)
fit_vis <- fit_vis[c(1, length(fit_vis))] # only keep best and worst fits

# best fits
fit_vis[[1]]

# worst fits
fit_vis[[2]]
```